Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-05 doc/invt.tex
code t0 = 2*c/a0, small changes

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
can often be improved
by the histogram along $i$ accumulated
during simulation
Using the histogram $H_i$ as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ H_i }
          { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
the previous value).
%
This scheme works well for early stages.
%
However, it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme.
%
Note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus, with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
In other words, the optimal $\alpha(t)$
is given by
\begin{equation}
\alpha(t)
=
\frac{    1     }
     {  t + t_0 }
,
\label{eq:alpha_invt1}
\end{equation}
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.

@@

Unfortunately,
Eq. \eqref{eq:q_opt} does not provide much information
on the behavior of $\alpha(t)$,
and the explicit solution
can be formidable in the general case.
%
We can nonetheless obtain an expression regarding $\alpha(t)$
by differentiating Eq. \eqref{eq:Lagrangian_const},
which yields
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  \equiv
  \lambda(t)
  .
  \label{eq:dinvadt}
\end{equation}
%
The expression in the middle is a weighted average
of the eigenvalues, $\lambda_k$'s,
and we shall refer to it as the instantaneous
eigenvalue, $\lambda(t)$.
%
The instantaneous eigenvalue
gives the rate of change of
the inverse updating magnitude.

@@

\end{tikzpicture}
\caption{
  \label{fig:vardep}
  Dependence of variables in the method of computing
  the optimal schedule.
}
\end{figure}

@@

To adapt the above formulae to a non-periodic variable,
we use the reflective boundary condition\cite{bussi2006},
and change the form of the updating matrix to
%
%
\begin{equation}
  w_{ij}
  =
  \mu_{ i - j }
  +
  \mu_{ i + j - 1 }
  +
  \mu_{ i + j - 2 n - 1 },
  \label{eq:w_band_refl}
\end{equation}
%
Again the updating kernel satisfies
Eqs. \eqref{eq:mu_normalization}
and \eqref{eq:mu_symm}.
%
However,
the radius $b$ only need to be less than $n$.
%
The last two terms
on the right-hand side of \eqref{eq:w_band_refl} are added
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006},
such that Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance}
are satisfied.

@@

\subsubsection{\label{sec:stabilize_wband}
Stabilization}

@@

The error of the equilibrium FDS method
depends on the precision of the histogram.
%
The error is minimized
when $v_i$'s assume the exact values,
%
and in this case,
the error came from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \log \frac { H_i }
                   { p_i }
      \right)^2
  \right\rangle
\notag
\\
  &\approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        - 1
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.

@@

In the long-time limit,
\begin{align*}
\left\langle
  \Theta_k^2
\right\rangle
&=
\frac{1}{T^2}
\sum_{ t, t' = 1 }^T
\langle \theta_k(t) \, \theta_k(t') \rangle
\\
&\approx
\frac{1}{T}
\sum_{ \tau = -\infty }^{ \infty }
\langle \theta_k(t) \, \theta_k(t + \tau) \rangle
\\
&=
\frac{1}{T}
\int_{-\infty}^\infty
\kappa_n(\tau) \, d\tau
=
\frac{1}{T}
\Gamma_k.
\end{align*}
%
Thus,
\begin{equation}
  E
  =
  \frac{ 1 } { T }
  \sum_{ k = 2 }^n \Gamma_k
  .
\label{eq:error_equil}
\end{equation}

@@

Again, the above argument can be extended
to the bandpass updating schemes.
%
If we are certain that
the intrinsic PMF is sufficiently smooth,
then we can explicitly filter some error modes,
say those for $k > K$,
in the histogram, and the sum in
Eq. \eqref{eq:error_equil}
can be truncated at $k = K$.

@@

We first tested the theory using
the inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a parameter to be optimized
with $t_0 = 2/(\lambda \, \alpha_0)$.
%
This simple schedule
allows us to predict the error analytically
[cf. Appendix \ref{sec:invt_schedule}],
and it serves as a model of verifying our theory.

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \eqref{sec:nnscheme}.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $T$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(T + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length, $T$,
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length, which is in agreement with
the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case,
the $T$-dependence was hardly visible
because of the predominantly large $\Gamma_k$'s
of the small $k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

While the optimal schedules of the Gaussian updating schemes
are rather complex,
those of the bandpass updating schemes
are always given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
with the optimal $t_0$
given by Eq. \eqref{eq:t0_sinc},
as shown in Fig. \ref{fig:sinc}(a).

In Fig. \ref{fig:sinc}(b),
we show that the updating kernels
of the bandpass updating schemes
are similar to Gaussian ones
except the oscillatory tails.
%
The equivalent width, $\sigma$, is roughly given
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  ,
\label{eq:sigma_equiv}
\end{equation}
where $g = 2$ for the periodic case
or $g = 1$ otherwise.
%
The kernel is a zigzag in the non-periodic case
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, the ruggedness disappears
in the actual updating matrix, Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by either the term for $l = i + j - 1$
or that for $l = i + j - 2 \, n - 1$,
as shown in the inset
of Fig. \ref{fig:sinc}(b).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    (a) Optimal schedules of the bandpass
    updating schemes for a periodic variable
    with $K = 4$,
    and for a non-periodic variable
    with $K = 8$.
    (b) Corresponding updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix according to Eq. \eqref{eq:w_band_refl}.
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
the optimal updating scheme
depends on the simulation length, $T$,
and other parameters.
%
To illustrate this,
we compare Gaussian updating schemes
with different width, $\sigma$.
%
For the global sampling process,
we see from Fig. \ref{fig:errsigscan}(a)
that the lowest error was achieved
by the widest window for the shorter simulation
of $T = 10^8$,
while the minimum moved to $\sigma = 0$
as the simulation lengthens to $T = 10^{10}$.
%
However,
the shift was hard to detect
for the local sampling scheme,
as shown in Fig. \ref{fig:errsigscan}(b).
%
In all cases,
the bandpass scheme
appeared to be a superior
alternative to the Gaussian updating scheme.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the brick-wall bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
  }
\end{center}
\end{figure}

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods
effectively sample a flat distribution
along a quantity of interest, $z$,
by constructing a bias potential along $z$
that offsets the PMF.
%
The optimal schedule ensures fastest convergence
of the bias potential to the PMF.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian wave packet.

The optimal schedule computed from our method
is identical to the inverse-time schedule,
Eq. \eqref{eq:alpha_invt1},
for the single-bin (WL) case.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule does not have a closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In the latter case,
the optimal schedule and error
are sensitive to the simulation length.

We have also compared different updating schemes.
%
We have shown that in the long time limit
the single-bin scheme belongs to a class of
brick-wall bandpass updating schemes that are
asymptotically optimal.
%
%That is, the single-bin scheme
%is asymptotically as efficient as
%an equilibrium FDS simulation
%with the exact bias potential.

@@

The optimal schedule predicted
by the direct solution of Eq. \eqref{eq:q_opt}
is complicated for a general
multiple-bin updating scheme.
%
Several previous studies\cite{marsili2006, barducci2008, dickson2011}
have been focused on analytically simpler schedules.
%
In a similar spirit,
we consider here
the inverse-time formula,
Eq. \eqref{eq:alpha_invtlambda},
%
with the proportionality constant, $\lambda$,
to be optimized.
%
We shall compute the error computed from
this variationally optimized schedule
and compare it with that from the optimal schedule.
%
We shall show that the optimal parameter, $\lambda$,
is sensitive to the simulation time, $T$,
which parallels the gradual shift of
the instantaneous eigenvalue,
$\lambda(t)$,
as discussed in Sec. \ref{sec:optschedule}.

@@

Using Eq. \eqref{eq:alpha_invtlambda}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { T + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        1 + \frac{ T }{ t_0 }
      \right)^{1 - 2 \, \nu_k}
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
At long times, $T \to \infty$, we get
$$
\begin{aligned}
  \Err_A
  =
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { T + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (T + t_0) }
    \ln \frac{ T + t_0 }
             {   t_0   }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (T + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases
are asymptotically suboptimal
as they decay slower than $1/T$.
%
Thus, by assuming exclusively the first case
$\lambda < 2 \lambda_k$,
the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_k$,
and can be identified as
the smallest real root of
%
\begin{equation}
\sum_{k = 2}^n
\frac{ \lambda_k - \lambda }
{ \left(2 - \lambda/ \lambda_k \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

Under the approximation, we have
$$
\frac{ \lambda_k^2 }
     { \lambda \, (2 \, \lambda_k - \lambda) }
\ge 1,
$$
with the equality achieved only at $\lambda = \lambda_k$.
For the equality to hold for all $k$'s,
we must have
$\lambda_2 = \dots = \lambda_n = \lambda$.
%
Now a matrix with uniform eigenvalues
is a multiple of the identity matrix,
which corresponds to the single-bin updating scheme.
%
The above result means that
the single-bin updating scheme is best
for the asymptotic convergence.
%
This result is discussed in more details in
Sec. \ref{sec:optWL}.

>>>>

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ H_i }
          { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with $\alpha$
reduced by a factor of one
half\cite{wang2001, wang2001pre}.
%
This scheme works well for early stages.
%
However, it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme.
%
Note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus, with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
In other words, the optimal $\alpha(t)$
is given by
\begin{equation}
\alpha(t)
=
\frac{    1     }
     {  t + t_0 }
,
\label{eq:alpha_invt1}
\end{equation}
and Eq. \eqref{eq:alpha_invt}
is the case of $t_0 = 0$.

@@

Unfortunately,
Eq. \eqref{eq:q_opt} does not provide much information
on the behavior of $\alpha(t)$,
and the explicit solution
can be formidable in the general case.
%
We can nonetheless obtain an expression regarding $\alpha(t)$
by differentiating Eq. \eqref{eq:Lagrangian_const},
which yields
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  \equiv
  \lambda(t)
  .
  \label{eq:dinvadt}
\end{equation}
%
The expression in the middle is a weighted average
of the eigenvalues, $\lambda_k$'s,
and we shall refer to it as the instantaneous
eigenvalue, $\lambda(t)$.
%
It gives the rate of change of
the inverse updating magnitude.

@@

\end{tikzpicture}
\caption{
  \label{fig:vardep}
  Dependence of variables in computing
  the optimal schedule.
}
\end{figure}

@@

To adapt the above formulae to a non-periodic variable,
we use the reflective boundary condition\cite{bussi2006},
and change the form of the updating matrix to
%
%
\begin{equation}
  w_{ij}
  =
  \mu_{ i - j }
  +
  \mu_{ i + j - 1 }
  +
  \mu_{ i + j - 2 n - 1 },
  \label{eq:w_band_refl}
\end{equation}
%
Again the updating kernel satisfies
Eqs. \eqref{eq:mu_normalization}
and \eqref{eq:mu_symm}.
%
However,
the cutoff $b$ only need to be less than $n$.
%
The last two terms
on the right-hand side of \eqref{eq:w_band_refl} are added
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006},
such that Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance}
are satisfied.

@@

\subsubsection{\label{sec:stabilize_wband}
Inversion and Stabilization}

@@

The error of the equilibrium FDS method
depends on the precision of the histogram.
%
The error is minimized
when $v_i$'s assume the exact values,
%
and in this case,
the error came from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \log \frac { H_i }
                   { p_i }
      \right)^2
  \right\rangle
%\notag
%\\
  %&
  \approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        - 1
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.

@@

In the long-time limit,
\begin{align*}
  \left\langle
    \Theta_k^2
  \right\rangle
  &=
  \frac{1}{T^2}
  \sum_{ t, t' = 1 }^T
  \langle \theta_k(t) \, \theta_k(t') \rangle
  %\\
  %&\approx
  %\frac{1}{T}
  %\sum_{ \tau = -\infty }^{ \infty }
  %\langle \theta_k(t) \, \theta_k(t + \tau) \rangle
  %\\
  %&=
  =
  \frac{1}{T}
  \int_{-\infty}^\infty
  \kappa_n(\tau) \, d\tau
  =
  \frac{ \Gamma_k }{ T }
.
\end{align*}
%
Thus,
\begin{equation}
  E
  =
  \frac{ 1 } { T }
  \sum_{ k = 2 }^n \Gamma_k
  .
\label{eq:error_equil}
\end{equation}

@@

Again, the above argument can be extended
to the bandpass updating schemes.
%
If we are certain that
the intrinsic PMF is sufficiently smooth,
then we can explicitly filter some error modes
in the histogram, and truncate the sum in
Eq. \eqref{eq:error_equil} at $k = K$.

@@

We first tested the theory using
the inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a parameter to be optimized
with $t_0 = 2/(\lambda \, \alpha_0)$.
%
The error of this schedule
can be computed analytically
(cf. Appendix \ref{sec:invt_schedule}),
and thus be tested against the theory.

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \eqref{sec:nnscheme}.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $T$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(T + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length, $T$,
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length,
which parallel the gradual shift of
the instantaneous eigenvalue $\lambda(t)$
as discussed in Sec. \ref{sec:optschedule}.
%which is in agreement with
%the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case,
the $T$-dependence was hardly visible
because of the predominantly large $\Gamma_k$'s
of the small $k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

On the other hand,
the optimal schedules of the bandpass updating schemes
are always given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
with the optimal $t_0$
given by Eq. \eqref{eq:t0_sinc},
as shown in Fig. \ref{fig:sinc}(a).

In Fig. \ref{fig:sinc}(b),
we show that the updating kernels
of the bandpass updating schemes
are similar to Gaussian ones
except the oscillatory tails.
%
The equivalent width, $\sigma$, is roughly given
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  ,
\label{eq:sigma_equiv}
\end{equation}
where $g = 2$ for the periodic case
or $g = 1$ otherwise.
%
The kernel is a zigzag in the non-periodic case
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, as shown in the inset of Fig. \ref{fig:sinc}(b),
the ruggedness disappears
in the actual updating matrix, Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by either the term for $l = i + j - 1$
or that for $l = i + j - 2 \, n - 1$,

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
the optimal updating scheme
depends on the simulation length, $T$,
and other parameters.
%
To illustrate this,
we compare Gaussian updating schemes
with different width, $\sigma$.
%
For the global sampling process,
we see from Fig. \ref{fig:errsigscan}(a)
that the minimal error was achieved
by the widest window for the shorter simulation
of $T = 10^8$,
while the minimum around $\sigma = 0$
became more favorable as the simulation lengthens.
%
However,
the shift was hard to detect
for the local sampling scheme,
as shown in Fig. \ref{fig:errsigscan}(b).

In contrast,
the normalized error of the bandpass updating scheme
was insensitive to the simulation length,
in agreement with Eq. \eqref{eq:error_sinc}.
%
Further, in all cases the bandpass scheme
appeared to be superior to the Gaussian updating scheme.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
  }
\end{center}
\end{figure}

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods
effectively sample a flat distribution
by constructing a bias potential
that offsets the PMF.
%
The optimal schedule ensures fastest convergence
of the bias potential.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian wave packet.

The optimal schedule computed from our method
recovers the known inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
for the single-bin (WL) case.
%
The origin of time, $t_0$, is better changed
to the value given by Eq. \eqref{eq:t0_sinc},
such that the initial updating magnitude
is half of the previous equilibrium value.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule does not have a closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
are sensitive to the simulation length.

We have also shown that in the long time limit
the single-bin scheme can be generalized
to a class of bandpass updating schemes that are
asymptotically optimal.
%
The bandpass updating schemes
eliminate local noise in the bias potential
without affecting the global convergence,
and they appeared to outperform
the Gaussian updating schemes.
%
Thus, the bandpass schemes
may be useful in real metadynamics simulations.
%
%That is, the single-bin scheme
%is asymptotically as efficient as
%an equilibrium FDS simulation
%with the exact bias potential.

@@

As an alternative to the
optimal schedule from Eq. \eqref{eq:q_opt},
%
here we consider the inverse-time formula,
Eq. \eqref{eq:alpha_invtlambda},
with $\lambda$ optimized variationally.
%
This approach allows to obtain
an explicit formula for the error,
which can be tested against simulation.
%
%Several previous studies\cite{marsili2006, barducci2008, dickson2011}
%have been focused on analytically simpler schedules.
%%
%In a similar spirit,
%we consider here
%the inverse-time formula,
%Eq. \eqref{eq:alpha_invtlambda},
%%
%%
%We shall compute the error computed from
%this variationally optimized schedule
%and compare it with that from the optimal schedule.
%%
%We shall show that the optimal parameter, $\lambda$,
%is sensitive to the simulation time, $T$,
%which parallels the gradual shift of
%the instantaneous eigenvalue,
%$\lambda(t)$,
%as discussed in Sec. \ref{sec:optschedule}.

@@

Using Eq. \eqref{eq:alpha_invtlambda}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { T + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        \frac {     t_0 }
              { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
The error is asymptotically minimal
only if $2\,\nu_k > 1$ for every $k$.
\note{At long times, $T \to \infty$, we get
$$
\begin{aligned}
  \Err_A
  =
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { T + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (T + t_0) }
    \ln \frac{ T + t_0 }
             {   t_0   }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (T + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases
are asymptotically suboptimal
as they decay slower than $1/T$.
}
%
In this case,
we can drop the second term
in the square brackets,
and identify the optimal $\lambda$ as
the smallest real root of
%
\begin{equation}
  \sum_{k = 2}^n
    \frac { \lambda_k - \lambda }
          { \left(2 - \lambda/ \lambda_k \right)^2 }
  = 0.
\notag
%\label{eq:optimal_lambda_approx}
\end{equation}
%
Since
$$
\frac{ \nu_k^2        }
     { 2 \, \nu_k - 1 }
=
\frac{ \lambda_k^2 }
     { \lambda \, (2 \, \lambda_k - \lambda) }
\ge 1,
$$
with the equality achieved only at $\lambda = \lambda_k$,
the optimal scheme must satisfy
$\lambda_2 = \dots = \lambda_n = \lambda$.
%
%Now a matrix with uniform eigenvalues
%is a multiple of the identity matrix,
%which corresponds to the single-bin updating scheme.
%
%The above result means that
%the single-bin updating scheme is best
%for the asymptotic convergence.
%
%This result is discussed in more details in
This again reflects the asymptotic optimality
of the single-bin scheme, as discussed in
Sec. \ref{sec:optWL}.

For a finite-length simulation,
we need to include the residual error, $\Err_R$.
%
By Eq. \eqref{eq:alpha_invtlambda}, we have
%
\begin{equation}
  q(T)
  =
  \frac{ 1 } { \lambda }
  \log\left(
    \frac{ T + t_0 } { t_0 }
  \right)
  ,
\notag
%\label{eq:qt_invtlambda}
\end{equation}
%
so that Eq. \eqref{eq:error_res1} becomes
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that
the system was initially equilibrated
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql},
as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      -
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.
